Archaic introgression and the distribution of shared functional variation under stabilizing selection

### 2023-09-28 doc/manuscript.tex
edits

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
        \aprcomment{$w$ and $f$ need to match text}
    }
    \label{fig:one-pop}
\end{figure}

@@

\subsection*{Theory and model}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration, and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{toy-admixture}A,B).

The second model is meant to more closely resemble inferred human-Neanderthal
history, in which the ancestral population of size $N_e=10\,000$ splits into
the human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
introgression contributes $5\%$ ancestry to Neanderthals. The human branch
shrinks to size $1\,000$ 60ka, followed by exponential growth to $20\,000$ at
present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked and
expanding human population at 50ka, after which their branch ceases
(Figure~\ref{fig:neand-to-human}A).

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is $0$ in all branches, and the strength of selection $V_S=1$
remains constant. \aprcomment{Justified by equivalence under rescaling
\citep{simons2018population}.}

@@

We expect genetic variance to increase after introgression. The amount that
genetic variance increases depends on the allelic differences accumulated
between populations and the effects of those alleles. Assuming no linkage,
dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
proportion $f$ contributed by the source population (labeled 0) into the focal
population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
expression for $V_G$, and after some simple algebra (Appendix~AX), we can
express the expected genetic variance directly after admixture as \[V_G =
fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\] where $F_2 = (p_0 -
p_1)^2$ is the squared difference in allele frequencies at a locus
\citep[e.g.,][]{peter2016admixture}. This result is known
\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
to that in the source populations weighted by their contributions, plus a term
that depends on the divergence at trait-affecting loci between the populations
weighted by the quadratic factor $2f(1-f)$.

In general, $F_2$ at a given locus will depend on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on the demography. In this case
\begin{align*}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] n_m V_M,
\end{align*}
where $V_M$ is the variance of effect sizes of new mutations (assuming the mean
is zero).

For $a \not\approx 0$, assuming neutral evolution will tend to overestimate
$F_2$ compared to underdominant selection. We can compute expected $V_G$ both
before and after admixture using the diffusion approximation for the joint SFS
with underdominance. Comparing to simulations with free recombination between
loci (Appendix~AY), we find that this provides an excellent fit to average
observed $V_G$ over time (Figure~\ref{fig:toy-admixture}C,D). Following an
initial rapid increase in $V_G$ due to admixture, it decays back to the
steady-state expectation relatively rapidly.

\begin{itemize}
    \item Benefit of SFS -- partition by frequency classes (prev. seg vs introgressed)
    \item VG from all both previously segregating and introgressed alleles decays,
        replaced by VG due to new mutations
    \item Dividing VG from introgressed alleles into introduced derived vs reintroduced
        ancestral alleles shows the effect of differences in population sizes
\end{itemize}

>>>>

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/one_pop.pdf}
    \caption{
        \textbf{Stabilizing selection and additive genetic variance in one population.}
        \aprcomment{See comparisons to simulations assuming free recombination in
        Figures~\ref{fig:one-popA}--\ref{fig:one-popC}.}
        \aprcomment{$w$ and $f$ need to match text}
        \aprcomment{Add the SHC model, and the ``vary SD'' fig as a panel here}
    }
    \label{fig:one-pop}
\end{figure}

@@

\subsection*{Model}

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration and admixture. Here, we
consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In a simple toy model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, one remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{fig:toy-admixture}A,B).

The second model is meant to more closely resemble inferred human-Neanderthal
history, in which the ancestral population of size $N_e=10\,000$ splits into
the human and Neanderthal branches. At 250ka, an early human-to-Neanderthal
introgression contributes $5\%$ ancestry to Neanderthals. The human branch
shrinks to size $1\,000$ 60ka, followed by exponential growth to $20\,000$ at
present time. Neanderthals contribute $2\%$ ancestry to this bottlenecked and
expanding human population at 50ka, after which they go extinct
(Figure~\ref{fig:neand-to-human}A).

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is kept at $0$ in all branches, and the strength of selection
$V_S=1$ remains constant.

@@

We expect genetic variance to increase after introgression. The amount that
genetic variance increases depends on the allelic differences accumulated
between populations and the effects of those alleles. Assuming no linkage,
dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
proportion $f$ contributed by the source population (labeled 0) into the focal
population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
expression for $V_G$, and after some simple algebra (Appendix~AX), we can
express the expected genetic variance directly after admixture as \[V_G =
fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\] where $F_2 = (p_0 -
p_1)^2$ is the squared difference in allele frequencies at a locus
\citep{peter2016admixture}. This result is known
\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
to that in the source populations weighted by their contributions, plus a term
that depends on the divergence at trait-affecting loci between the populations
weighted by the quadratic factor $2f(1-f)$.

In general, $F_2$ at a given locus will depend on the demographic history
relating the two populations and the effect size at the locus due to selection
on the trait. In the infinitesimal limit, involving many loci each of
vanishingly small effect, dynamics at a given locus will be approximately
neutral, so that $F_2$ depends only on the demography. In this case
\begin{align*}
    V_G & \approx f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] \sum_l a_l^2 \\
    & = f V_{G,0} + (1-f) V_{G,1} + 2f(1-f)\mathbb{E}[F_2] L V_M,
\end{align*}
where $V_M$ is the variance of effect sizes of new mutations (assuming the mean
is zero), and $L$ is the number of trait affecting loci.

For $a \not\approx 0$, the assumption of neutral evolution will lead to
overestimates of $F_2$ compared to underdominant selection. We can compute
expected $V_G$ both before and after admixture using the diffusion
approximation for the joint SFS with underdominance. Comparing to simulations
with free recombination between loci (Appendix~AY), we find that this provides
an excellent fit to average observed $V_G$ over time
(Figure~\ref{fig:toy-admixture}C,D). In general, admixture causes a sudden
increase in $V_G$ followed by a fairly rapid return to pre-admixture levels.

Modeling the dynamics of genetic variance using the SFS lets us examine
contributions to $V_G$ from different classes of mutations, such as those at
different frequencies or arising at different times, and how those
contributions change over time. Of interest is the contribution to $V_G$ from
mutations that were already segregating in the recipient population, those that
were introduced through introgression, and new mutations since the time of
admixture.

Paragraph on the following (point to appendix and supplement):
\begin{itemize}
    \item Benefit of SFS -- partition by frequency classes (prev. seg vs introgressed)
    \item VG from all both previously segregating and introgressed alleles decays,
        replaced by VG due to new mutations
    \item Dividing VG from introgressed alleles into introduced derived vs reintroduced
        ancestral alleles shows the effect of differences in population sizes
\end{itemize}
